---
title: "Understanding Time Series in R"
output:
  github_document: default
  html_notebook: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(forecast)
library(broom)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center"
)
```

## Objective

The objectives are:

* To use the modelling tools in the `forecast` package.

* To understand various packages available within R for working with time series e.g. base, `zoo`, `xts`.

* To understand how the above can be integrated into an analysis which uses the `tidyverse` family of packages.

## Simulate Data

Imagine that we intend to explore a time series collected each month over 10 years starting from 2005.  It will be assumed that the time series data has a trend, a seasonal component and an error.  The time series can be simulated as follows:

```{r}
# for reproducibility ------------------------------------------------------
set.seed(20170214) 

# sample times -------------------------------------------------------------
start_date <- ymd(20050101)
end_date <- ymd(20141201)
date <- seq(from = start_date, to = end_date, by = "month")
         
# data components ----------------------------------------------------------
level <- 1
trend <- (row_number(date) - 1) / 12
seasonal <- sin(2 * pi * row_number(date) / 12) 
error <- rnorm(length(date), sd = 0.5)
value <- level + trend + seasonal + error

# timeseries as data frame -------------------------------------------------
ts_df <- data_frame(date, value)

# plot timeseries ----------------------------------------------------------
g_ts <- ggplot(ts_df, aes(date, value)) +
    geom_line()
```

In the above code, we constructed an object of `dataframe` class i.e.

```{r}
class(ts_df)
```

This is desirable since the powerful `tidyverse` can now be put to use on it.  Unfortunately though, specialist modelling tools that are applicable to time series analysis often require objects which are `ts` or `ts` derivative classes.  Specifically, we are interested in using the `forecast` package and it is the case that objects should be `ts` classes to work with the `forecast` package.

## Base R

It is useful to understand how a `ts` class is constructed.  Consider the base R function `stats::stl`.  According to the documentation, this is a function for "decomposing a time series into seasonal, trend and irregular components".  Hence, it is a useful tool to apply to data similar to the s documentation for the `stats::stl` function states that it operates on and object which is:

* A univariate time series

* A `ts` class (i.e. time series)

* Has a frequency greater one.

It must be noted that the `stl` function cannot *guess* the period of seasonality and this must be explicitly defined when defining the frequency time series.  Consider the following `ts` objects (we use `value` from the simulation defined above):

```{r}
ts(value[1:10])
```

According to the documentation, a `ts` object is restricted to be "sampled at equidistant points in time".  By default, the frequency is assumed to be 1; the  frequency is defined as the number of observations per unit time.  Note that when frequencies of 4 and 12 are used, they are special cases and the series are assumed to be quarterly and yearly respectively: 

```{r}
ts(value[1:12], frequency = 4)
ts(value[1:36], frequency = 12)
```

A start time may be supplied as "a single number or a vector of two integers, which specify a natural time unit and a (1-based) number of samples into the time unit".  For example, to start 4 samples from the beginning of 2010:

```{r}
ts(value, start = c(2010, 4), frequency = 12)
```

It is convenient to work with time series to gain access to packages that provide modelling features for time series.  A time series is constructed as follows:

```{r}
raw_ts <- ts(value, start = 2015, frequency = 12)
raw_ts
```

Frankly, `ts` objects not very intuitive and they do no behave consistently i.e. the special cases.  It is typical of base R: when used at the console, they provide 'convenience' features but when used programmatically, the inconsistencies can cause grief. In the current case, the simulation can be constructed as a base `ts` object as follows: 
 
```{r}
(ts_base <- ts(value, start = year(start_date), freq = 12))
```

And we can perform a STL decomposition:

```{r}
fit <- stl(ts_base, s.window = "periodic")
autoplot(fit)
```

It is important to emphasise that the programmer must explicitly fix the period of the season (in this case, annually) through the `frequency` argument when constructing the `ts`.  For example, if the above decomposition is repeated but with `frequency` set to 4 (for quarterly samples), the result is different (and clearly wrong) even though the "core data" is the same: 

```{r}
fit <- stl(ts(value, start = 2005, frequency = 4), s.window = "periodic")
autoplot(fit)
```

## Using `ts` together with `tidyverse`

The aim here will be to use `forecast::ma` to compute a moving average and to superpose it onto a plot of the raw data using `ggplot`.  First, the moving average is computed as follows:
   
```{r}
mod_ma <- ma(ts_base, order = 5)
```

The command `broom::tidy` can be used to transform this to a data frame:

```{r}
ts_to_df <- function(ts, index) {
    vals <- ts %>% 
        tidy() %>%
        select_(setNames(1, "value")) %>% 
        mutate(value = as.numeric(value))
    cbind(index, vals) 
}

ma_df <- ts_to_df(mod_ma, date)
```

This can be plotted as with `ggplot`:

```{r}
g_ts + geom_line(data = ma_df, aes(date, value), na.rm = TRUE, 
                 col = 'blue', size = 0.3)
```
